---
title: "Differences in Fuel Efficiency Between Automatic and Manual Transmission Vehicles in 1974"
author: "Wayne Witzke"
header-includes:
  - \usepackage[table]{xcolor}
output: pdf_document
---

## Synopsis
We attempt to establish whether an automatic or manual transmission provides
better fuel efficiency, and quantify that fuel savings if a difference exists,
using vehicle data collected by Motor Trend magazine for 1973-1974 models. A
minimal adequate linear regression model is fitted against the data, including
transmission type as a factor variable to attempt to isolate the transmission
contribution to fuel efficiency. No significant contribution is found.

## Preliminaries

This section prepares the analysis to run. For more information on replicating
this analysis, see the appendix.

\scriptsize
```{r setup}
library(ggplot2); library(GGally);
data(mtcars);
options(digits=4, width=125);
```
\normalsize

## Exploratory Analysis

### Summary of Data

This analysis uses the `mtcars` data set included natively with most R
distributions. This data includes 11 aspects of automobile design and
performance for 32 different models of automobile from 1973 and 1974. These
aspects include: miles per gallon (`mpg`); the number of cylinders (`cyl`,
either 4, 6, or 8); the displacement (`disp`, in cubic inches); the gross
horsepower (`hp`); the rear axle ratio (`drat`); the weight (`wt`, in 1000's of
pounds); the quarter-mile time (`qsec`); the engine type (`vs`, either v-engine
or straight engine); the transmission type (`am`, automatic or manual); the
number of gears (`gear`, either 3, 4 or 5); and the number of carburetors
(`carb`, either 1, 2, 3, 4, 6, or 8). The structure of the data set is:

\scriptsize
```{r data_structure, results="hold"}
dim(mtcars);
sapply(mtcars,class);
```
\normalsize

### Exploratory Graphs

We are primarily interested in characterizing the relationship between miles
per gallon and transmission type. However, there is correlation between these
variables and other variables in the dataset. We can see this with a pairs
plot.  We can get a better overall characterization of the data by tidying it,
specifically by ensuring that factor variables are properly expressed.

\scriptsize
```{r tidying_data}
mtc = mtcars; mtc$cyl = factor(mtc$cyl); mtc$gear = factor(mtc$gear);
mtc$carb = factor(mtc$carb);
mtc$am = factor(mtc$am, labels = c("auto", "man"));
mtc$vs = factor(mtc$vs, labels = c("v", "str"));
```
\normalsize

Figure 1 in the appendix shows the pairs plot with this corrected data, using
box plots and faceted density plots to show relationships between continuous
and discrete variables, and faceted bar plots to show relationships between
factors.

From this pair plot, we can see that fitting a simple, single-variable,
meaningful linear regression is likely just not possible. That is, `mpg`
appears to correlate with many of the variables available. In addition, it
appears very likely that there are confounding variables. For instance, the
pairs `mpg`/`wt`, `mpg`/`disp`, and `wt`/`disp` are all strongly correlated.

## Regression Modeling

Because of the strong correlations between `mpg` and the other variables, and
because of the strong possibility of confounding, we use a backward selection
strategy with `anova` to attempt to find a multivariable regression model that
isolates the impact of transmission on mpg. This involves removing variables
from a complete model until we have found the minimum adequate fit, using both
F-test p-values and Akaike information criterion (`AIC`) for selection.
Interactions are also tested. The details of and code for this procedure can be
found in Figure 2 in the appendix.

Once run, the procedure selects `mpg ~ am + wt + hp + cyl` as the best model.
The details of this model, including coefficients, confidence intervals,
leverage and influence tests, can be seen here.

\scriptsize
```{r show_regression_model, results="hold"}
mtc.fit = lm(mpg ~ am + wt + hp + cyl, mtc);
summary(mtc.fit)$coef;
confint(mtc.fit);
head(hatvalues(mtc.fit)[order(hatvalues(mtc.fit),decreasing=TRUE)],6);
head(dffits(mtc.fit)[order(abs(dffits(mtc.fit)),decreasing=TRUE)],6);
```
\normalsize

Residual disgnostic plots can be seen in Figure 3 in the appendix.

## Conclusions

The quality of the best linear regression fit is fairly good. The residual
plots (Figure 3) do not show any marked deviations from what might be expected
from a good model, and examining the highest leverage and influence points does
not reveal problematic outliers. Most of the p-values from the model satisfy
$\alpha=0.05$, indicating that they are likely significant contributing
variables in the analysis. The coefficients seem reasonable as well. That is,
it is reasonable that for every 1000 pound increase in weight, you lose about
2.5 miles per gallon, or that as horsepower increases by 1, you lose about 0.03
miles per gallon.

Unfortunately, the transmission type was *not* one of the coefficients that
appeared to contribute significantly to the model. At $p\approx0.2$, it fails
to reject the null hypothesis that a model including transmission type is
identical to a model that does not include it. Either there is no contribution
from transmission or there is not enough data to detect an existing significant
contribution. This makes quantifying any such contribution impossible with this
data set.


\newpage
## Appendix
\scriptsize

### Figure 1: Pairs Plot
```{r tidy_pairs_plot, fig.width = 8, fig.height = 7}
tpplot = ggpairs(mtc[, c("qsec","drat","hp","disp","wt","mpg","am","cyl","vs","gear","carb")],
    lower=list(continuous=wrap("smooth", size=0.2),combo="facetdensity",discrete="blank"),
    upper=list(continuous=wrap("cor", size=2.5, color="black"),combo="box",discrete="facetbar"), axisLabels="none");
suppressMessages(print(tpplot, left = 0.75, bottom = 0.75));
```

### Figure 2: Backward Selection + Anova to Find Minimal Adequate Fit
```{r linear_regression_start, results="hold"}
fit = list(lm(mpg ~ am+wt+disp+hp+drat+qsec+cyl+vs+gear+carb, mtc)); # Start with full model.
fit[[2]] = update(fit[[1]], ~.-wt); fit[[3]] = update(fit[[1]], ~.-disp); fit[[4]] = update(fit[[1]], ~.-hp);
fit[[5]] = update(fit[[1]], ~.-drat); fit[[6]] = update(fit[[1]], ~.-qsec); fit[[7]] = update(fit[[1]], ~.-cyl);
fit[[8]] = update(fit[[1]], ~.-vs); fit[[9]] = update(fit[[1]], ~.-gear); fit[[10]] = update(fit[[1]], ~.-carb);
round1 = rbind(sapply(fit[1:10], function(n) anova(fit[[1]],n)[,"Pr(>F)"][2]),sapply(fit[1:10], AIC));
colnames(round1) = c("base","wt","disp","hp","drat","qsec","cyl","vs","gear","carb");
round1[, order(round1[2,])]; # Select "carb" to remove for high p-value and low AIC.
fit = list(lm(mpg ~ am+wt+disp+hp+drat+qsec+cyl+vs+gear, mtc)); # Start with full model.
fit[[2]] = update(fit[[1]], ~.-wt); fit[[3]] = update(fit[[1]], ~.-disp); fit[[4]] = update(fit[[1]], ~.-hp);
fit[[5]] = update(fit[[1]], ~.-drat); fit[[6]] = update(fit[[1]], ~.-qsec); fit[[7]] = update(fit[[1]], ~.-cyl);
fit[[8]] = update(fit[[1]], ~.-vs); fit[[9]] = update(fit[[1]], ~.-gear);
round2 = rbind(sapply(fit[1:9], function(n) anova(fit[[1]],n)[,"Pr(>F)"][2]),sapply(fit[1:9], AIC));
colnames(round2) = c("base","wt","disp","hp","drat","qsec","cyl","vs","gear");
round2[, order(round2[2,])]; # Select "gear" to remove for high p-value and low AIC.
fit = list(lm(mpg ~ am+wt+disp+hp+drat+qsec+cyl+vs, mtc)); # Start with full model.
fit[[2]] = update(fit[[1]], ~.-wt); fit[[3]] = update(fit[[1]], ~.-disp); fit[[4]] = update(fit[[1]], ~.-hp);
fit[[5]] = update(fit[[1]], ~.-drat); fit[[6]] = update(fit[[1]], ~.-qsec); fit[[7]] = update(fit[[1]], ~.-cyl);
fit[[8]] = update(fit[[1]], ~.-vs);
round2 = rbind(sapply(fit[1:8], function(n) anova(fit[[1]],n)[,"Pr(>F)"][2]),sapply(fit[1:8], AIC));
colnames(round2) = c("base","wt","disp","hp","drat","qsec","cyl","vs");
round2[, order(round2[2,])]; # Select "drat" to remove for high p-value and low AIC.
fit = list(lm(mpg ~ am+wt+disp+hp+qsec+cyl+vs, mtc)); # Start with full model.
fit[[2]] = update(fit[[1]], ~.-wt); fit[[3]] = update(fit[[1]], ~.-disp); fit[[4]] = update(fit[[1]], ~.-hp);
fit[[5]] = update(fit[[1]], ~.-qsec); fit[[6]] = update(fit[[1]], ~.-cyl); fit[[7]] = update(fit[[1]], ~.-vs);
round2 = rbind(sapply(fit[1:7], function(n) anova(fit[[1]],n)[,"Pr(>F)"][2]),sapply(fit[1:7], AIC));
colnames(round2) = c("base","wt","disp","hp","qsec","cyl","vs");
round2[, order(round2[2,])]; # Select "disp" to remove for high p-value and low AIC.
fit = list(lm(mpg ~ am+wt+hp+qsec+cyl+vs, mtc)); # Start with full model.
fit[[2]] = update(fit[[1]], ~.-wt); fit[[3]] = update(fit[[1]], ~.-hp); fit[[4]] = update(fit[[1]], ~.-qsec);
fit[[5]] = update(fit[[1]], ~.-cyl); fit[[6]] = update(fit[[1]], ~.-vs);
round2 = rbind(sapply(fit[1:6], function(n) anova(fit[[1]],n)[,"Pr(>F)"][2]),sapply(fit[1:6], AIC));
colnames(round2) = c("base","wt","hp","qsec","cyl","vs");
round2[, order(round2[2,])]; # Select "qsec" to remove for high p-value and low AIC.
fit = list(lm(mpg ~ am+wt+hp+cyl+vs, mtc)); # Start with full model.
fit[[2]] = update(fit[[1]], ~.-wt); fit[[3]] = update(fit[[1]], ~.-hp); fit[[4]] = update(fit[[1]], ~.-cyl);
fit[[5]] = update(fit[[1]], ~.-vs);
round2 = rbind(sapply(fit[1:5], function(n) anova(fit[[1]],n)[,"Pr(>F)"][2]),sapply(fit[1:5], AIC));
colnames(round2) = c("base","wt","hp","cyl","vs");
round2[, order(round2[2,])]; # Select "vs" to remove for high p-value and low AIC.
fit = list(lm(mpg ~ am+wt+hp+cyl, mtc)); # Start with full model.
fit[[2]] = update(fit[[1]], ~.-wt); fit[[3]] = update(fit[[1]], ~.-hp); fit[[4]] = update(fit[[1]], ~.-cyl);
round2 = rbind(sapply(fit[1:4], function(n) anova(fit[[1]],n)[,"Pr(>F)"][2]),sapply(fit[1:4], AIC));
colnames(round2) = c("base","wt","hp","cyl");
round2[, order(round2[2,])]; # Additional removals will make the fit worse. So, done.
fit = lm(mpg ~ am + wt + hp + cyl, mtc);
fitI2 = update(fit, ~.^2);
fitI3 = update(fit, ~.^3);
fitI4 = update(fit, ~.^4);
anova(fit, fitI2)[,"Pr(>F)"][2]; # Checking for interactions (i.e. "am:wt").
anova(fit, fitI3)[,"Pr(>F)"][2]; # More interactions.
anova(fit, fitI4)[,"Pr(>F)"][2]; # Last interaction ("am:wt:hp:cyl"). Note no significant difference when including interactions.


```

### Figure 3: Residual Diagnostics
```{r residual_diagnostics, fig.height=6}
par(mfrow=c(2,2));
plot(mtc.fit);
par(mfrow=c(1,1));
```

### Figure 4: System Information

This analysis was performed using the hardware and software specified in this
section.

\scriptsize
```{r system_info}
sessionInfo();
```
